% Current working manuscript with available experiment values filled in.
% Values marked \tbd require experiments that have not been run yet.

\documentclass[pdflatex,sn-mathphys-num]{sn-jnl}

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
\usepackage{tikz}%
\usetikzlibrary{arrows.meta,positioning,shapes.geometric,fit}%

\newcommand{\tbd}{\textsc{TBD}}

\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}%

\raggedbottom

\begin{document}

\title[Dynamic Trust Calibration]{Dynamic Trust Calibration: Multi-Agent Contextual Arbitration for Mitigating Bias in Retina Foundation Models}

\author*[1,2]{\fnm{Abdul Mohaimen Al} \sur{Radi}}\email{iauthor@gmail.com}
\author[2]{\fnm{Yusra} \sur{Ahmed}}\email{iiauthor@gmail.com}
\equalcont{These authors contributed equally to this work.}
\author[2]{\fnm{Nguyen} \sur{Bui}}\email{iiiauthor@gmail.com}
\equalcont{These authors contributed equally to this work.}
\author[1,2]{\fnm{Yu} \sur{Tian}}\email{iiiauthor@gmail.com}

\affil*[1]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{100190}, \state{State}, \country{Country}}}
\affil[2]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{10587}, \state{State}, \country{Country}}}
\affil[3]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{610101}, \state{State}, \country{Country}}}

\abstract{Retinal foundation models can achieve strong average performance while remaining unreliable for specific diseases, imaging modalities, or patient subgroups. This mismatch between aggregate accuracy and patient-level reliability is a central barrier to equitable clinical deployment. We introduce Equi-Agent, a multi-agent framework that performs dynamic trust calibration for retinal disease assessment. Rather than averaging model outputs or applying a fixed post-hoc threshold, Equi-Agent arbitrates among modality-specific foundation models using patient metadata, validation-derived subgroup reliability priors, visual morphology review, functional visual-field evidence, and a final safety audit. We evaluate this approach on Harvard-FairVision for cross-sectional diagnosis of age-related macular degeneration, diabetic retinopathy, and glaucoma, and extend the experimental framework to Harvard-GDP for longitudinal glaucoma progression forecasting. Across controlled transfer experiments, standalone retinal foundation models show substantial subgroup-specific false-positive, false-negative, and calibration gaps despite reasonable aggregate performance. These findings motivate contextual arbitration as a distinct layer between foundation-model inference and clinical decision support. Equi-Agent provides an interpretable mechanism for converting model disagreement and subgroup reliability evidence into calibrated predictions, uncertainty estimates, and escalation decisions, offering a reproducible framework for studying equitable deployment of ophthalmic foundation models.}

\keywords{ophthalmic foundation models, fairness, calibration, multi-agent systems, retina imaging}

\maketitle

\section{Introduction}
Retinal diseases such as age-related macular degeneration (AMD), diabetic retinopathy (DR), and glaucoma remain major causes of preventable visual impairment. Their diagnosis depends on subtle structural and functional evidence distributed across fundus photography, scanning laser ophthalmoscopy (SLO), optical coherence tomography (OCT), and visual-field testing. Automated systems are therefore attractive for scalable eye care, particularly in settings where specialist access is limited \cite{TODO}.

The recent emergence of ophthalmic foundation models has changed the technical landscape. Large pretrained retinal encoders can be adapted to multiple downstream tasks and often outperform smaller task-specific models. Yet average accuracy is an incomplete measure of clinical readiness. A model may perform well overall while being poorly calibrated for a subgroup, overcalling disease in one demographic stratum, or missing early pathology in another. This failure mode is especially consequential in ophthalmology, where anatomical baselines and disease appearance vary across age, ancestry, image modality, acquisition protocol, and disease stage.

Current mitigation strategies only partially address this problem. Static ensembles assume that model reliability can be averaged. Calibration methods rescale probabilities without deciding which evidence source should be trusted. Fairness-aware post-processing adjusts thresholds at the group level, but often treats subgroup membership as a fixed correction rather than as one component of a broader clinical context. In practice, the relevant question is more granular: given this patient, this image, this disease task, and this pattern of model disagreement, which output should influence the final decision, and with what degree of confidence?

Here we introduce Equi-Agent, a multi-agent framework for contextual arbitration among retinal foundation models. Equi-Agent treats model trust as dynamic rather than intrinsic. It integrates foundation-model predictions with patient metadata, validation-derived subgroup reliability priors, morphology-grounded visual review, visual-field interpretation, guideline context, and a final safety audit. The framework is designed to produce not only a diagnostic label, but also a calibrated confidence estimate, a transparent account of model trust, and an escalation recommendation when evidence is discordant or insufficient.

We make three contributions. First, we establish a controlled benchmark for comparing retinal foundation models under identical FairVision train, validation, and test splits with standardized subgroup and calibration metrics. Second, we formulate dynamic trust calibration as an explicit arbitration problem, separating visual evidence, subgroup reliability, model disagreement, functional context, and safety review into distinct agents. Third, we show that existing foundation models and non-agentic fusion baselines leave substantial subgroup reliability gaps, motivating an agentic layer between foundation-model inference and clinical decision support.

\section{Related Work}
\subsection{Foundation Models in Ophthalmic Imaging}
\label{subsec:related_foundation}
Foundation models for ophthalmic imaging learn reusable representations from large-scale OCT, fundus, SLO, or multimodal image-text corpora. RETFound represents OCT-oriented self-supervised pretraining, whereas MIRAGE, FLAIR, RET-CLIP, VisionFM, UrFound, and RetiZero reflect a broader family of retinal encoders intended for transfer across downstream tasks \cite{TODO}. These systems differ substantially in modality support, pretraining objective, image resolution, and downstream adaptation recipe, making uncontrolled comparison difficult.

We therefore evaluate foundation models under a shared transfer protocol wherever possible. OCT-oriented models are evaluated on OCT inputs, fundus/SLO-oriented models on SLO/fundus inputs, and all downstream thresholds are selected using validation data only. This protocol compares transferability under a common evaluation contract. It is not intended to be a claim that every foundation model has been tuned to its model-specific optimum.

\subsection{Bias and Fairness in Medical Vision Models}
Medical AI systems can reproduce or amplify disparities when training data, acquisition protocols, disease prevalence, or label quality differ across patient groups \cite{TODO}. In ophthalmology, these concerns include unequal sensitivity, unequal false-positive burden, and subgroup-specific calibration error. Such failures are not merely statistical; they can alter referral patterns, delay treatment, or increase unnecessary follow-up for already underserved populations.

Equi-Agent addresses fairness as a problem of reliability-aware inference. Rather than using demographic metadata as disease evidence, the framework uses validation-set behavior to estimate when a model's output is less trustworthy for a given subgroup and task. This distinction is central: demographic context informs trust in the model, not the biological likelihood of disease.

\subsection{Multi-Agent and LLM-Based Clinical Reasoning Systems}
Large language models have been used to summarize clinical records, retrieve guidelines, explain model outputs, and support multi-step medical reasoning \cite{TODO}. Multi-agent systems extend this idea by assigning specialized roles to separate reasoning modules. This can improve interpretability, but it also creates risks: hallucinated findings, prompt sensitivity, and poorly controlled decision logic.

Equi-Agent constrains language-model reasoning to evidence-grounded arbitration. Agents reason over supplied model probabilities, patient metadata, validation priors, image-derived findings, visual-field measurements, and guideline summaries. They are not used to infer hidden clinical measurements or replace the image models. The empirical question is whether this structured arbitration improves reliability beyond deterministic routing, static fusion, and post-hoc calibration.

\section{Methods}
\subsection{Problem Formulation}
\label{subsec:problem_formulation}
Let $x_i^{(m)}$ denote the imaging input for patient-eye case $i$ and modality $m$, and let $a_i$ denote available patient metadata, including age group, race, ethnicity, and sex/gender. For each disease task $t \in \{\mathrm{AMD}, \mathrm{DR}, \mathrm{glaucoma}\}$, a set of base models $\mathcal{M}$ produces probabilities $p_{ij}^{(t)} = f_j(x_i^{(m_j)})$ for model $j$. The objective is to estimate a final prediction $\hat{y}_i^{(t)}$ and confidence score that preserve aggregate diagnostic performance while reducing subgroup-specific reliability gaps.

Equi-Agent frames this as contextual arbitration. For a case $i$, the system estimates the trust assigned to model $j$ as a function of disease task, modality, patient metadata, subgroup-specific validation reliability, model confidence, and cross-model disagreement:
\begin{equation}
w_{ij}^{(t)} = g\left(t, m_j, a_i, p_{ij}^{(t)}, \mathcal{R}_{j,t,a_i}, \Delta_i^{(t)}\right),
\end{equation}
where $\mathcal{R}_{j,t,a_i}$ denotes validation-derived reliability priors for the relevant subgroup context and $\Delta_i^{(t)}$ denotes disagreement among available model outputs. The final prediction is produced by an arbitration procedure that combines weighted model evidence with independent visual, functional, and safety assessments. This formulation separates two quantities that are often conflated: the probability of disease and the reliability of the model producing that probability.

\subsection{Datasets and Prediction Tasks}
\label{subsec:methods_datasets}
Harvard-FairVision is used for cross-sectional diagnosis of AMD, DR, and glaucoma. Each task is represented with fixed train, validation, and test splits, and each prediction row is linked to demographic metadata used for subgroup evaluation. Harvard-GDP is used for glaucoma detection and progression forecasting, including OCT B-scan, retinal nerve fiber layer thickness (RNFLT), visual-field, and clinical metadata fields where available. FairVision supports the primary cross-sectional fairness analysis, whereas Harvard-GDP tests whether arbitration can be extended to temporal disease progression.

\subsection{Base Vision Foundation Models}
\label{subsec:base_models_methods}
We evaluate modality-specific foundation-model outputs using a standardized prediction schema. RETFound is evaluated as an OCT model and MIRAGE as an SLO/fundus model. FLAIR, RET-CLIP, and RetiZero are evaluated as SLO/fundus image encoders with task-specific linear probes. VisionFM and UrFound are evaluated with both Fundus/SLO and OCT inputs. For FLAIR, RET-CLIP, RetiZero, VisionFM, and UrFound, image embeddings are extracted from the frozen pretrained model and a balanced logistic regression head is trained for each FairVision disease task. Thresholds are selected on the validation split using the prespecified validation metric for each run, and final metrics are computed only on the held-out test split.

This design separates representation evaluation from model-specific fine-tuning optimization. It allows controlled comparison under the same downstream split, label, threshold, and metric contract, while acknowledging that some foundation models may perform differently under native full fine-tuning or prompt-based protocols.

\subsection{Equi-Agent Architecture}
\label{subsec:architecture_methods}
Equi-Agent is organized as a structured arbitration system (Fig.~\ref{fig:equi_agent_architecture}). The BioProfiler converts structured metadata into a concise clinical narrative. The Vision Specialist processes OCT and SLO/fundus inputs, reports foundation-model probability distributions, and performs a morphology-grounded visual audit. The Functional Specialist interprets visual-field information, especially mean deviation (MD), to provide glaucoma-relevant functional context. The Equity Auditor compares the current patient context with validation-derived subgroup error priors and recommends whether sensitivity- or precision-oriented trust adjustment is warranted. The Guidelines Agent retrieves and summarizes external clinical guidance when available. The Orchestrator integrates these findings into final disease labels. The Safety Agent performs a final uncertainty and discordance audit and can recommend acceptance, override, escalation to human review, or insufficient-data status.

\begin{figure*}[t]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
    font=\small,
    node distance=8mm and 10mm,
    box/.style={draw, rounded corners=2pt, align=center, minimum width=27mm, minimum height=10mm, fill=gray!5},
    agent/.style={draw, rounded corners=2pt, align=center, minimum width=31mm, minimum height=11mm, fill=blue!5},
    data/.style={draw, align=center, minimum width=27mm, minimum height=9mm, fill=green!5},
    outputbox/.style={draw, rounded corners=2pt, align=center, minimum width=35mm, minimum height=11mm, fill=orange!8},
    arrow/.style={-{Latex[length=2mm]}, thick}
]
\node[data] (image) {OCT / SLO\\images};
\node[data, below=of image] (metadata) {Patient metadata\\and visual field};
\node[agent, right=14mm of image] (vision) {Vision\\Specialist};
\node[agent, right=14mm of metadata] (bio) {BioProfiler};
\node[agent, below=of bio] (functional) {Functional\\Specialist};
\node[box, above=of vision] (models) {Foundation models\\RETFound, MIRAGE,\\FLAIR, RET-CLIP,\\VisionFM, RetiZero, UrFound};
\node[box, below=of functional] (priors) {Validation-derived\\subgroup priors};
\node[agent, right=17mm of bio] (equity) {Equity\\Auditor};
\node[agent, above=of equity] (guidelines) {Guidelines\\Agent};
\node[agent, right=17mm of equity] (orchestrator) {Final\\Orchestrator};
\node[agent, below=of orchestrator] (safety) {Safety\\Agent};
\node[outputbox, right=16mm of orchestrator] (output) {Disease labels,\\confidence, rationale,\\escalation flag};

\draw[arrow] (image) -- (vision);
\draw[arrow] (models) -- (vision);
\draw[arrow] (metadata) -- (bio);
\draw[arrow] (metadata) |- (functional);
\draw[arrow] (vision) -- (equity);
\draw[arrow] (bio) -- (equity);
\draw[arrow] (priors) -| (equity);
\draw[arrow] (bio) |- (guidelines);
\draw[arrow] (vision) -- (orchestrator);
\draw[arrow] (functional) -| (orchestrator);
\draw[arrow] (equity) -- (orchestrator);
\draw[arrow] (guidelines) -- (orchestrator);
\draw[arrow] (orchestrator) -- (safety);
\draw[arrow] (safety) -- (output);
\draw[arrow] (orchestrator) -- (output);
\end{tikzpicture}
}
\caption{Equi-Agent architecture. The system separates image-derived evidence, clinical metadata, subgroup reliability priors, functional visual-field interpretation, guideline context, final arbitration, and safety review into explicit agents.}
\label{fig:equi_agent_architecture}
\end{figure*}

\subsection{Dynamic Trust Calibration}
\label{subsec:dynamic_trust_methods}
Dynamic trust calibration uses validation-set behavior to estimate when each model is more or less reliable. For each base model, disease task, and subgroup attribute, we compute validation AUROC, F1-score, false-positive rate, false-negative rate, expected calibration error, and sample counts. Subgroup priors are marked unstable when positive or negative counts are too small. At inference time, the arbitration layer first checks stable intersectional priors, then single-attribute priors, then global model priors.

The Equity Auditor converts these priors into trust recommendations. A high subgroup false-negative rate motivates a sensitivity shift, lowering the effective diagnostic threshold or increasing uncertainty when model confidence is borderline. A high false-positive rate motivates a precision shift, raising the threshold or down-weighting overconfident positive predictions. These adjustments act on model trust and confidence; demographic metadata is not used as direct disease evidence.

\subsection{Contextual Arbitration Procedure}
\label{subsec:arbitration_methods}
The arbitration procedure receives structured inputs: model probabilities, validation thresholds, subgroup priors, patient metadata, visual morphology summaries, functional summaries, and guideline context. The Orchestrator applies disease-specific rules and agent summaries to produce final labels. For glaucoma, functional evidence can constrain image-only predictions when the visual field is normal; in the current implementation, a mean deviation greater than $-2$ dB is treated as evidence against glaucoma unless image evidence is compelling. For AMD, the system uses a pathology-signal representation derived from cumulative stage probabilities and inspects whether disease probability mass is concentrated in early, intermediate, or advanced stages. For DR, arbitration emphasizes vascular lesion evidence and cross-model confidence.

The Safety Agent then audits the decision for discordant structural and functional evidence, missing modalities, poor image quality, or unusually high uncertainty. This final stage is designed to prevent inappropriate reassurance and to identify cases that should be escalated rather than forced into a binary output.

\subsection{Longitudinal Progression Reasoning}
\label{subsec:longitudinal_methods}
For Harvard-GDP, longitudinal extension treats each visit sequence as a temporal evidence stream. The temporal agent compares current and prior OCT/RNFL evidence, visual-field change, and model confidence trajectories. A progression decision should be supported by clinically plausible temporal change rather than by a single noisy scan. GDP baselines use RNFLT, OCT B-scan features, clinical metadata, and their combinations, and provide the non-agentic reference point for longitudinal arbitration.

\subsection{Evaluation Metrics}
\label{subsec:metrics_methods}
Evaluation is implemented with a standardized prediction schema containing patient identifiers, task, model name, split, ground-truth label, predicted probability, thresholded prediction, and demographic metadata. For final agentic comparisons, we emphasize thresholded diagnostic metrics, including F1-score, balanced accuracy, sensitivity, specificity, Brier score, negative log-likelihood, and expected calibration error. Ranking metrics such as AUROC are retained only for standalone non-agentic model diagnostics where probability ranking is directly interpretable. Subgroup evaluation includes per-subgroup performance, worst-group F1, false-positive rate gap, false-negative rate gap, equal opportunity difference, equalized odds difference, subgroup ECE, and intersectional subgroup metrics.

\subsection{Statistical Analysis}
\label{subsec:statistical_methods}
Unless otherwise stated, metrics are computed on the held-out test split after thresholds are selected on validation data. Confidence intervals will be estimated by nonparametric patient-level bootstrapping in the final analysis. Paired comparisons will be made on shared test cases, with multiple subgroup comparisons controlled using the Benjamini--Hochberg procedure.

\section{Experimental Design and Evaluation}\label{sec:experiments}

The experimental evaluation tests whether dynamic trust calibration improves the reliability of ophthalmic foundation models without sacrificing aggregate diagnostic performance. The central hypothesis is that model reliability is not fixed globally, but varies as a function of disease task, imaging modality, patient metadata, subgroup-specific error patterns, model disagreement, and longitudinal context. Equi-Agent is therefore evaluated against standalone foundation models, static ensembles, post-hoc calibration methods, rule-based arbitration, and fairness-aware baselines.

The experiments address four questions. First, do standalone ophthalmic foundation models exhibit subgroup-specific reliability gaps despite strong aggregate performance? Second, does Equi-Agent improve diagnostic performance and worst-group reliability compared with standalone models and conventional ensembles? Third, is agentic arbitration necessary, or can deterministic routing and static fusion provide the same benefit? Fourth, does the same arbitration framework extend to longitudinal glaucoma progression prediction?

Demographic metadata is used only as contextual information for estimating model reliability under subgroup-specific failure modes. It is not used as a direct proxy for disease status. Metadata-only, metadata-shuffled, and no-demographic ablations are included to evaluate this design choice.

To ensure controlled comparison among foundation models, all models are evaluated using the same FairVision train, validation, and test splits, validation-only threshold selection, and held-out test metrics. Each model is evaluated on the imaging modality most consistent with its intended use. The protocol compares transferability under a common downstream setting rather than exhaustively optimizing each model-specific recipe.

\subsection{Datasets and Prediction Tasks}\label{subsec:datasets}

We evaluate Equi-Agent on ophthalmic diagnostic and longitudinal prediction tasks using Harvard-FairVision and Harvard-GDP. Harvard-FairVision is used for cross-sectional eye disease diagnosis and subgroup fairness evaluation. Harvard-GDP is used for longitudinal glaucoma progression prediction. For each dataset, we record imaging modality, disease labels, demographic attributes, train/validation/test splits, and subgroup sample sizes.

\begin{table*}[t]
\centering
\caption{Dataset and task summary. Demographic attributes and subgroup sample sizes are reported to support fairness and worst-group reliability analyses.}
\label{tab:dataset_summary}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lllllllll@{}}
\toprule
Dataset & Modality & Task & Disease labels & Patients / eyes & Images / visits & Demographic attributes & Split & Primary use \\
\midrule
Harvard-FairVision & OCT B-scans / SLO fundus & Cross-sectional diagnosis & AMD, DR, glaucoma & \tbd & 10,000 examples per task; 30,000 task-label rows & Race, ethnicity, sex/gender, age & 6,000 train / 1,000 val / 3,000 test per task & Diagnostic fairness \\
Harvard-GDP & OCT B-scan / RNFLT & Glaucoma detection and progression forecasting & Glaucoma, progression & \tbd & 1,000 glaucoma-detection examples; 500 progression examples used & Race, sex/gender, age, Hispanic ethnicity & Detection: 600 train / 400 test; progression: 300 train / 200 test & Longitudinal reasoning \\
External validation cohort, if available & \tbd & Cross-dataset generalization & \tbd & \tbd & \tbd & \tbd & Held-out & Robustness \\
\botrule
\end{tabular}%
}
\end{table*}

\subsection{Base Models and Baselines}\label{subsec:baselines}

Equi-Agent is compared against three baseline families. The first includes standalone ophthalmic and general-purpose vision foundation models, such as RETFound, MIRAGE, FLAIR, RET-CLIP, VisionFM, RetiZero, UrFound, DINOv2, CLIP-style retinal vision-language models, and task-specific supervised CNN or Transformer models. The second includes static fusion methods, including majority vote, mean probability ensembling, confidence-weighted ensembling, and logistic stacking. The third includes fairness and calibration baselines, including group reweighting, group distributionally robust optimization, subgroup temperature scaling, group-specific thresholding, and equalized-odds post-processing.

Where code and training details are available, we also compare against ophthalmology-specific fairness methods, including Fair Identity Scaling, FairerOPTH-style implicit fairness learning, and longitudinal glaucoma fairness baselines such as FairDist or EqEffNet-style models. If a baseline cannot be reproduced exactly, we report the closest implementable variant and describe the difference explicitly.

\begin{table*}[t]
\centering
\caption{Baseline families used to evaluate whether Equi-Agent provides benefits beyond standalone foundation models, post-hoc calibration, subgroup-aware routing, and static ensembling.}
\label{tab:baseline_summary}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lllll@{}}
\toprule
Baseline family & Method & Uses demographic metadata? & Uses model disagreement? & Purpose \\
\midrule
Foundation models & RETFound OCT & No & No & Standalone OCT foundation-model baseline for FairVision diagnosis \\
Foundation models & MIRAGE SLO & No & No & Standalone fundus/SLO foundation-model baseline for FairVision diagnosis \\
Foundation models & FLAIR SLO & No & No & Standalone fundus/SLO vision-language foundation-model baseline using FairVision linear probing \\
Foundation models & RET-CLIP SLO & No & No & Standalone fundus/SLO vision-language foundation-model baseline using FairVision linear probing \\
Foundation models & VisionFM SLO & No & No & Standalone VisionFM fundus/SLO encoder baseline using FairVision linear probing \\
Foundation models & VisionFM OCT & No & No & Standalone VisionFM OCT encoder baseline using FairVision linear probing \\
Foundation models & RetiZero SLO & No & No & Standalone RetiZero fundus/SLO vision-language encoder baseline using FairVision linear probing \\
Foundation models & UrFound SLO & No & No & Standalone UrFound fundus/SLO encoder baseline using FairVision linear probing \\
Foundation models & UrFound OCT & No & No & Standalone UrFound OCT encoder baseline using FairVision linear probing \\
Static ensembles & Mean probability ensemble & No & Weak / implicit & Test whether simple averaging of RETFound OCT and MIRAGE SLO is sufficient \\
Static ensembles & Confidence-weighted ensemble & No & Weak / implicit & Test whether confidence-weighted model fusion improves calibration and diagnostic reliability \\
Calibration / thresholding & Validation-selected task thresholds & No & No & Correct task-specific probability scale mismatch using validation-only threshold tuning \\
Subgroup reliability priors & Validation-derived subgroup priors & Yes & No & Estimate subgroup-specific model reliability, false-positive patterns, false-negative patterns, and calibration risk \\
Rule-based arbitration & Dynamic global prior routing by validation ranking & No & Yes & Select the globally most reliable model per disease task using validation performance \\
Rule-based arbitration & Dynamic global prior routing by ECE & No & Yes & Select the most calibrated model or ensemble per disease task using validation ECE \\
Rule-based arbitration & Dynamic subgroup prior routing & Yes & Yes & Test whether subgroup-specific validation priors can improve per-patient model selection \\
Agent baselines & Single-agent LLM arbitration & Optional & Yes & Test whether a single reasoning agent can improve over deterministic routing \\
Agent baselines & Multi-agent without subgroup priors & Optional & Yes & Test whether model disagreement reasoning helps without explicit fairness priors \\
Equi-Agent & Multi-agent contextual arbitration with subgroup priors & Yes & Yes & Final proposed framework for dynamic trust calibration and fairness-aware diagnostic arbitration \\
\botrule
\end{tabular}%
}
\end{table*}

\subsection{Equi-Agent Arbitration Framework}\label{subsec:equi_agent_framework}

Equi-Agent receives patient imaging data, patient metadata, predictions from multiple base models, model confidence scores, localized visual evidence, and validation-derived subgroup error priors. The system is structured as a multi-agent arbitration framework. A disease evidence agent evaluates whether the visual findings support the candidate diagnosis. A demographic context agent assesses whether patient metadata changes the reliability of specific visual cues or model outputs. A bias auditor compares the current prediction context against subgroup-specific false-positive and false-negative patterns estimated from the validation set. A model disagreement agent identifies conflicts among base models and determines whether disagreement reflects uncertainty, modality-specific weakness, or subgroup-specific unreliability. For longitudinal tasks, a temporal reasoning agent evaluates whether changes across visits are clinically plausible. A final arbitration agent integrates these outputs into a calibrated prediction, confidence score, and explanation.

To reduce stochasticity and improve reproducibility, agent outputs are constrained to a structured schema. Each agent must return a fixed set of fields, including the evidence considered, the reliability concern identified, the recommended trust adjustment, and a confidence estimate. The arbitration agent is instructed to use only the provided model outputs, metadata, subgroup priors, and image-derived findings.

\subsection{Evaluation Metrics}\label{subsec:metrics}

We report aggregate diagnostic performance and subgroup-specific reliability. Aggregate performance is measured using F1-score, balanced accuracy, sensitivity, specificity, and calibration metrics. Ranking metrics are used only for standalone model diagnostics where appropriate. Fairness and subgroup reliability are measured using worst-group F1, false-positive rate gap, false-negative rate gap, equal opportunity difference, equalized odds difference, subgroup expected calibration error, and subgroup Brier score.

For a demographic attribute $A$ with subgroups $g \in \mathcal{G}$, the false-positive rate gap is defined as
\begin{equation}
\Delta \mathrm{FPR} = \max_{g \in \mathcal{G}} \mathrm{FPR}_{g} - \min_{g \in \mathcal{G}} \mathrm{FPR}_{g}.
\label{eq:fpr_gap}
\end{equation}

The false-negative rate gap is defined as
\begin{equation}
\Delta \mathrm{FNR} = \max_{g \in \mathcal{G}} \mathrm{FNR}_{g} - \min_{g \in \mathcal{G}} \mathrm{FNR}_{g}.
\label{eq:fnr_gap}
\end{equation}

Worst-group F1 is defined as
\begin{equation}
\mathrm{F1}_{\mathrm{worst}} = \min_{g \in \mathcal{G}} \mathrm{F1}_{g}.
\label{eq:worst_group_f1}
\end{equation}

For calibration, we report expected calibration error both globally and within each subgroup. Subgroup ECE is computed by binning predictions within each subgroup and measuring the weighted difference between confidence and empirical accuracy. We also report reliability diagrams, Brier score, and risk-coverage curves when abstention or uncertainty-aware decision-making is evaluated.

\subsection{Experiment 1: Reliability Gaps in Standalone Foundation Models}\label{subsec:exp1}

The first experiment establishes the failure mode motivating Equi-Agent. Each standalone base model is evaluated on each disease task, with aggregate diagnostic performance reported alongside subgroup-specific error disparities. This experiment tests whether strong average performance masks clinically meaningful reliability gaps across demographic groups.

\begin{table*}[t]
\centering
\caption{Standalone ophthalmic models can exhibit subgroup-specific reliability gaps despite strong aggregate performance. $\Delta$FPR and $\Delta$FNR denote max-min subgroup disparity for false-positive and false-negative rates. Metrics are reported on the Harvard-FairVision held-out test split after validation-selected task thresholds.}
\label{tab:standalone_failure_modes}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}llllllllll@{}}
\toprule
Model & Dataset & Disease & F1 $\uparrow$ & Sens. $\uparrow$ & Spec. $\uparrow$ & Worst-group F1 $\uparrow$ & $\Delta$FPR $\downarrow$ & $\Delta$FNR $\downarrow$ & Subgroup ECE $\downarrow$ \\
\midrule
RETFound OCT & Harvard-FairVision & AMD & 0.775 & 0.772 & 0.878 & 0.286 & 0.208 & 0.502 & 0.120 \\
MIRAGE SLO & Harvard-FairVision & AMD & 0.638 & 0.762 & 0.652 & 0.200 & 0.124 & 0.233 & 0.239 \\
FLAIR SLO & Harvard-FairVision & AMD & 0.565 & 0.771 & 0.469 & 0.133 & 0.228 & 0.240 & 0.623 \\
RET-CLIP SLO & Harvard-FairVision & AMD & 0.611 & 0.740 & 0.623 & 0.190 & 0.243 & 0.405 & 0.599 \\
VisionFM SLO & Harvard-FairVision & AMD & 0.629 & 0.724 & 0.680 & 0.190 & 0.154 & 0.362 & 0.315 \\
VisionFM OCT & Harvard-FairVision & AMD & 0.743 & 0.786 & 0.817 & 0.279 & 0.225 & 0.379 & 0.264 \\
RetiZero SLO & Harvard-FairVision & AMD & 0.633 & 0.659 & 0.764 & 0.180 & 0.111 & 0.345 & 0.346 \\
UrFound SLO & Harvard-FairVision & AMD & 0.596 & 0.554 & 0.832 & 0.246 & 0.113 & 0.325 & 0.384 \\
UrFound OCT & Harvard-FairVision & AMD & 0.752 & 0.704 & 0.907 & 0.244 & 0.144 & 0.499 & 0.233 \\
\midrule
RETFound OCT & Harvard-FairVision & DR & 0.537 & 0.785 & 0.891 & 0.402 & 0.129 & 0.110 & 0.111 \\
MIRAGE SLO & Harvard-FairVision & DR & 0.316 & 0.762 & 0.708 & 0.222 & 0.287 & 0.140 & 0.148 \\
FLAIR SLO & Harvard-FairVision & DR & 0.247 & 0.766 & 0.576 & 0.148 & 0.319 & 0.168 & 0.393 \\
RET-CLIP SLO & Harvard-FairVision & DR & 0.282 & 0.785 & 0.639 & 0.181 & 0.372 & 0.288 & 0.384 \\
VisionFM SLO & Harvard-FairVision & DR & 0.299 & 0.579 & 0.782 & 0.208 & 0.154 & 0.255 & 0.188 \\
VisionFM OCT & Harvard-FairVision & DR & 0.440 & 0.774 & 0.834 & 0.347 & 0.214 & 0.138 & 0.203 \\
RetiZero SLO & Harvard-FairVision & DR & 0.283 & 0.743 & 0.666 & 0.184 & 0.193 & 0.247 & 0.238 \\
UrFound SLO & Harvard-FairVision & DR & 0.258 & 0.690 & 0.652 & 0.162 & 0.317 & 0.133 & 0.250 \\
UrFound OCT & Harvard-FairVision & DR & 0.480 & 0.747 & 0.870 & 0.360 & 0.106 & 0.218 & 0.133 \\
\midrule
RETFound OCT & Harvard-FairVision & Glaucoma & 0.721 & 0.720 & 0.737 & 0.478 & 0.490 & 0.494 & 0.183 \\
MIRAGE SLO & Harvard-FairVision & Glaucoma & 0.660 & 0.644 & 0.707 & 0.382 & 0.363 & 0.504 & 0.320 \\
FLAIR SLO & Harvard-FairVision & Glaucoma & 0.572 & 0.495 & 0.776 & 0.333 & 0.244 & 0.366 & 0.146 \\
RET-CLIP SLO & Harvard-FairVision & Glaucoma & 0.654 & 0.636 & 0.707 & 0.450 & 0.304 & 0.412 & 0.152 \\
VisionFM SLO & Harvard-FairVision & Glaucoma & 0.657 & 0.679 & 0.628 & 0.506 & 0.265 & 0.271 & 0.271 \\
VisionFM OCT & Harvard-FairVision & Glaucoma & 0.668 & 0.593 & 0.826 & 0.364 & 0.258 & 0.388 & 0.272 \\
RetiZero SLO & Harvard-FairVision & Glaucoma & 0.663 & 0.658 & 0.689 & 0.429 & 0.281 & 0.361 & 0.198 \\
UrFound SLO & Harvard-FairVision & Glaucoma & 0.690 & 0.778 & 0.544 & 0.495 & 0.389 & 0.343 & 0.231 \\
UrFound OCT & Harvard-FairVision & Glaucoma & 0.694 & 0.650 & 0.787 & 0.490 & 0.280 & 0.320 & 0.159 \\
\botrule
\end{tabular}%
}
\end{table*}

\subsection{Experiment 2: Main Diagnostic Comparison}\label{subsec:exp2}

The second experiment evaluates whether Equi-Agent improves diagnostic performance and subgroup reliability compared with standalone models, static ensembles, calibration methods, and fairness-aware baselines. The primary comparison is conducted on the held-out test set. Hyperparameters, calibration parameters, stacking weights, subgroup priors, and arbitration rules are selected using validation data only.

\begin{table*}[t]
\centering
\caption{Main diagnostic comparison against standalone foundation models, static ensembles, calibration methods, and rule-based dynamic trust calibration. Metrics are macro-averaged across AMD, DR, and glaucoma on the Harvard-FairVision held-out test split. Entries marked \tbd indicate outputs or subgroup-metric files that were not available in the local manuscript snapshot.}
\label{tab:main_diagnostic_comparison}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lllllllllll@{}}
\toprule
Method & Type & F1 $\uparrow$ & Balanced Acc. $\uparrow$ & Sens. $\uparrow$ & Spec. $\uparrow$ & Worst-group F1 $\uparrow$ & $\Delta$FPR $\downarrow$ & $\Delta$FNR $\downarrow$ & ECE $\downarrow$ & Subgroup ECE $\downarrow$ \\
\midrule
RETFound OCT & Foundation model & 0.678 & 0.797 & 0.759 & 0.835 & 0.286 & 0.490 & 0.502 & 0.054 & 0.183 \\
MIRAGE SLO & Foundation model & 0.538 & 0.706 & 0.723 & 0.689 & 0.200 & 0.363 & 0.504 & 0.103 & 0.320 \\
FLAIR SLO & Foundation model & 0.461 & 0.642 & 0.677 & 0.607 & 0.133 & 0.319 & 0.366 & 0.247 & 0.623 \\
RET-CLIP SLO & Foundation model & 0.516 & 0.688 & 0.720 & 0.656 & 0.181 & 0.372 & 0.412 & 0.245 & 0.599 \\
VisionFM SLO & Foundation model & 0.528 & 0.679 & 0.661 & 0.697 & 0.190 & 0.265 & 0.362 & 0.175 & 0.315 \\
VisionFM OCT & Foundation model & 0.617 & 0.772 & 0.718 & 0.826 & 0.279 & 0.258 & 0.388 & 0.137 & 0.272 \\
RetiZero SLO & Foundation model & 0.526 & 0.697 & 0.687 & 0.707 & 0.180 & 0.281 & 0.361 & 0.156 & 0.346 \\
UrFound SLO & Foundation model & 0.515 & 0.675 & 0.674 & 0.676 & 0.162 & 0.389 & 0.343 & 0.162 & 0.384 \\
UrFound OCT & Foundation model & 0.642 & 0.777 & 0.701 & 0.854 & 0.244 & 0.280 & 0.499 & 0.107 & 0.233 \\
Mean probability ensemble & Static ensemble & 0.469 & 0.577 & 0.730 & 0.425 & 0.058 & 0.509 & 0.484 & 0.081 & 0.311 \\
Confidence-weighted ensemble & Static ensemble & 0.476 & 0.584 & 0.829 & 0.339 & 0.108 & 0.509 & 0.484 & 0.077 & 0.369 \\
Dynamic global prior by validation ranking & Rule-based arbitration & 0.473 & 0.587 & 0.683 & 0.490 & 0.067 & 0.490 & 0.498 & 0.184 & 0.636 \\
Dynamic global prior by ECE & Rule-based calibration & 0.476 & 0.585 & 0.829 & 0.340 & 0.108 & 0.509 & 0.484 & 0.060 & 0.311 \\
Dynamic subgroup prior by validation ranking & Subgroup rule-based arbitration & 0.474 & 0.585 & 0.713 & 0.457 & 0.084 & 0.490 & 0.498 & 0.182 & 0.608 \\
\textbf{Equi-Agent} & \textbf{Dynamic arbitration} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} \\
\botrule
\end{tabular}%
}
\end{table*}

\begin{table*}[t]
\centering
\caption{Interim live Equi-Agent arbitration results on a stratified Harvard-FairVision subset. This placeholder table reports the completed 1,500-case live Azure run, with 500 cases per task and 250 positives / 250 negatives per task. Results are shown to document the current operating point before full-test inference and should be replaced or relabeled after the final 9,000 task-case run.}
\label{tab:equi_agent_stratified1500_placeholder}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lllllllllllll@{}}
\toprule
Task & $n$ & Pos./Neg. & F1 $\uparrow$ & Acc. $\uparrow$ & Balanced Acc. $\uparrow$ & Precision $\uparrow$ & Sens. $\uparrow$ & Spec. $\uparrow$ & FPR $\downarrow$ & FNR $\downarrow$ & Escalation rate & Threshold mix \\
\midrule
AMD & 500 & 250 / 250 & 0.486 & 0.496 & 0.496 & 0.496 & 0.476 & 0.516 & 0.484 & 0.524 & 0.584 & 0.35: 135; 0.50: 251; 0.65: 114 \\
DR & 500 & 250 / 250 & 0.404 & 0.522 & 0.522 & 0.536 & 0.324 & 0.720 & 0.280 & 0.676 & 0.912 & 0.35: 174; 0.50: 302; 0.65: 24 \\
Glaucoma & 500 & 250 / 250 & 0.785 & 0.790 & 0.790 & 0.803 & 0.768 & 0.812 & 0.188 & 0.232 & 0.578 & 0.35: 11; 0.50: 243; 0.65: 246 \\
Macro-average & 1500 & 750 / 750 & 0.558 & 0.603 & 0.603 & 0.612 & 0.523 & 0.683 & 0.317 & 0.477 & 0.691 & -- \\
\botrule
\end{tabular}%
}
\end{table*}

\begin{table*}[t]
\centering
\caption{Exact-subset comparison for the interim stratified 1,500-case live run. All rows are evaluated on the same 500 AMD, 500 DR, and 500 glaucoma cases used by the live Equi-Agent run. Some standalone models achieve high macro F1 by operating at very high sensitivity and very low specificity for AMD or DR; balanced accuracy, specificity, and calibration are therefore reported alongside F1.}
\label{tab:equi_agent_stratified1500_comparison}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lllllllllll@{}}
\toprule
Method & Type & F1 $\uparrow$ & Balanced Acc. $\uparrow$ & Sens. $\uparrow$ & Spec. $\uparrow$ & ECE $\downarrow$ & Worst-group F1 $\uparrow$ & $\Delta$FPR $\downarrow$ & $\Delta$FNR $\downarrow$ & Interpretation \\
\midrule
RETFound OCT & Single model & 0.674 & 0.607 & 0.805 & 0.408 & 0.326 & 0.298 & 0.294 & 0.365 & High sensitivity; weak specificity \\
RET-CLIP SLO & Single model & 0.672 & 0.532 & 0.955 & 0.109 & 0.042 & 0.416 & 0.220 & 0.142 & Near-all-positive AMD/DR behavior \\
FLAIR SLO & Single model & 0.669 & 0.521 & 0.968 & 0.075 & 0.045 & 0.401 & 0.141 & 0.105 & Near-all-positive AMD/DR behavior \\
Dynamic global validation-ranking source & Deterministic routing & 0.597 & 0.595 & 0.683 & 0.508 & 0.133 & 0.158 & 0.295 & 0.382 & More balanced than high-recall baselines \\
\textbf{Equi-Agent live} & \textbf{LLM arbitration} & \textbf{0.558} & \textbf{0.603} & \textbf{0.523} & \textbf{0.683} & \textbf{0.115} & \textbf{0.209} & \textbf{0.345} & \textbf{0.466} & \textbf{Conservative on AMD/DR} \\
Mean probability ensemble & Static fusion & 0.373 & 0.599 & 0.337 & 0.861 & 0.135 & 0.095 & 0.270 & 0.397 & High specificity; misses positives \\
Confidence-weighted ensemble & Static fusion & 0.342 & 0.594 & 0.309 & 0.879 & 0.172 & 0.095 & 0.262 & 0.364 & High specificity; low sensitivity \\
Validation ranking-weighted ensemble & Static fusion & 0.377 & 0.601 & 0.337 & 0.865 & 0.134 & 0.095 & 0.271 & 0.397 & Similar to mean fusion \\
\botrule
\end{tabular}%
}
\end{table*}

\subsection{Experiment 3: Fairness Across Demographic and Intersectional Subgroups}\label{subsec:exp3}

The third experiment evaluates whether Equi-Agent reduces error disparities across demographic attributes and intersectional subgroups. We report subgroup performance for race, ethnicity, sex/gender, age, and clinically relevant intersections such as race-by-sex and race-by-age. Intersectional results are reported only when subgroup sample sizes are sufficient for reliable estimation.

\begin{table*}[t]
\centering
\caption{Subgroup error disparities before and after dynamic trust calibration. Baseline refers to the strongest non-agentic comparator selected on validation performance.}
\label{tab:subgroup_fairness}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lllllllll@{}}
\toprule
Attribute & Subgroup comparison & Baseline F1 & Equi-Agent F1 & Baseline $\Delta$FPR & Equi-Agent $\Delta$FPR & Baseline $\Delta$FNR & Equi-Agent $\Delta$FNR & Worst-group F1 gain \\
\midrule
Race & Max-min subgroup gap & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
Ethnicity & Max-min subgroup gap & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
Sex / gender & Max-min subgroup gap & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
Age & Max-min subgroup gap & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
Race $\times$ sex/gender & Worst intersectional subgroup & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
Race $\times$ age & Worst intersectional subgroup & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
Sex/gender $\times$ age & Worst intersectional subgroup & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
Disease $\times$ subgroup & Worst disease-subgroup pair & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
\botrule
\end{tabular}%
}
\end{table*}

\subsection{Experiment 4: Dynamic Arbitration Versus Static Fusion}\label{subsec:exp4}

The fourth experiment tests whether dynamic trust calibration provides benefits beyond static model fusion. We compare Equi-Agent against majority vote, mean probability ensembling, confidence-weighted ensembling, logistic stacking, and group-specific thresholding. This experiment directly tests whether the proposed agentic arbitration layer is necessary.

\begin{table*}[t]
\centering
\caption{Dynamic trust calibration versus static model fusion on Harvard-FairVision. Aggregate metrics are macro-averaged across AMD, DR, and glaucoma.}
\label{tab:dynamic_vs_static}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}llllllllll@{}}
\toprule
Method & Uses confidence? & Uses subgroup priors? & Uses disagreement? & Uses metadata? & F1 $\uparrow$ & Worst-group F1 $\uparrow$ & $\Delta$FPR $\downarrow$ & $\Delta$FNR $\downarrow$ & ECE $\downarrow$ \\
\midrule
Mean probability ensemble & Yes & No & No & No & 0.469 & 0.058 & 0.509 & 0.484 & 0.081 \\
Confidence-weighted ensemble & Yes & No & Weak & No & 0.476 & 0.108 & 0.509 & 0.484 & 0.077 \\
Rule-based arbitration, global validation prior & Yes & No & Yes & No & 0.473 & 0.067 & 0.490 & 0.498 & 0.184 \\
Rule-based arbitration, global ECE prior & Yes & No & Yes & No & 0.476 & 0.108 & 0.509 & 0.484 & 0.060 \\
Rule-based arbitration, subgroup validation prior & Yes & Yes & Yes & Yes & 0.474 & 0.084 & 0.490 & 0.498 & 0.182 \\
\textbf{Equi-Agent} & \textbf{Yes} & \textbf{Yes} & \textbf{Yes} & \textbf{Yes} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} \\
\botrule
\end{tabular}%
}
\end{table*}

\subsection{Experiment 5: Ablation Study}\label{subsec:exp5}

The fifth experiment isolates the contribution of each Equi-Agent component. We remove one component at a time and measure changes in aggregate performance, worst-group reliability, fairness gaps, and calibration. This experiment tests whether demographic context, subgroup priors, model disagreement reasoning, localized visual evidence, and multi-agent structure contribute meaningfully to the final prediction.

\begin{table*}[t]
\centering
\caption{Ablation study of Equi-Agent components. Each row removes or modifies one component of the full framework.}
\label{tab:ablation}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lllllllll@{}}
\toprule
Variant & Removed / modified component & F1 $\uparrow$ & Balanced Acc. $\uparrow$ & Worst-group F1 $\uparrow$ & $\Delta$FPR $\downarrow$ & $\Delta$FNR $\downarrow$ & ECE $\downarrow$ & Brier $\downarrow$ \\
\midrule
\textbf{Full Equi-Agent} & None & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} \\
No demographic context agent & Patient metadata context removed & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
No bias auditor & Subgroup error priors removed & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
No model disagreement agent & Cross-model conflict reasoning removed & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
No visual evidence agent & Localized visual findings removed & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
No calibration agent & Confidence recalibration removed & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
Single-agent arbitration & Multi-agent structure collapsed into one agent & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
Rule-based arbitration & LLM arbitration replaced with deterministic rules & 0.473 & 0.587 & \tbd & \tbd & \tbd & 0.184 & \tbd \\
Randomized priors & Subgroup error priors randomly permuted & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
\botrule
\end{tabular}%
}
\end{table*}

\subsection{Experiment 6: Metadata Safety and Counterfactual Sensitivity}\label{subsec:exp6}

Because Equi-Agent uses demographic metadata as context for trust calibration, we evaluate whether this metadata is being used appropriately. The goal is to verify that metadata does not function as a direct disease shortcut. We compare image-only models, metadata-only models, naive image-plus-metadata models, Equi-Agent without metadata, Equi-Agent with shuffled metadata, and full Equi-Agent.

We also perform counterfactual metadata sensitivity tests. In these tests, image data and base-model outputs are held fixed while demographic metadata is changed. A robust arbitration system should not casually flip disease predictions solely because metadata changes. Instead, demographic context should primarily affect model trust weights, uncertainty estimates, or escalation decisions when subgroup-specific validation evidence supports such adjustment.

\begin{table*}[t]
\centering
\caption{Metadata safety and counterfactual sensitivity analysis. Demographic metadata should improve trust calibration without acting as a direct proxy for disease status.}
\label{tab:metadata_safety}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}llllllll@{}}
\toprule
Variant / test & Image input & Metadata input & Subgroup priors & F1 $\uparrow$ & Worst-group F1 $\uparrow$ & $\Delta$FPR $\downarrow$ & $\Delta$FNR $\downarrow$ \\
\midrule
Image-only classifier & Yes & No & No & \tbd & \tbd & \tbd & \tbd \\
Metadata-only classifier & No & Yes & No & \tbd & \tbd & \tbd & \tbd \\
Naive image + metadata classifier & Yes & Yes & No & \tbd & \tbd & \tbd & \tbd \\
Equi-Agent without metadata & Yes & No & No / partial & \tbd & \tbd & \tbd & \tbd \\
Equi-Agent with shuffled metadata & Yes & Shuffled & Yes & \tbd & \tbd & \tbd & \tbd \\
Equi-Agent with randomized priors & Yes & Yes & Randomized & \tbd & \tbd & \tbd & \tbd \\
\textbf{Full Equi-Agent} & \textbf{Yes} & \textbf{Yes} & \textbf{Validation-derived} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} \\
\botrule
\end{tabular}%
}
\end{table*}

\begin{table*}[t]
\centering
\caption{Counterfactual metadata sensitivity tests. Prediction instability is measured as the percentage of cases in which the final disease label changes when metadata is altered while image data and base-model outputs remain fixed.}
\label{tab:counterfactual_metadata}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lllllll@{}}
\toprule
Counterfactual test & Held fixed & Modified variable & Prediction flip rate $\downarrow$ & Mean confidence change & Trust-weight change & Desired behavior \\
\midrule
Race counterfactual & Image, base predictions & Race label & \tbd & \tbd & \tbd & Low label instability; bounded trust adjustment \\
Sex/gender counterfactual & Image, base predictions & Sex/gender label & \tbd & \tbd & \tbd & Low label instability; bounded trust adjustment \\
Age counterfactual & Image, base predictions & Age group & \tbd & \tbd & \tbd & Clinically plausible confidence adjustment \\
Metadata shuffling & Image, base predictions & All metadata & \tbd & \tbd & \tbd & Reduced fairness benefit versus full model \\
Prior randomization & Image, base predictions, metadata & Subgroup priors & \tbd & \tbd & \tbd & Reduced fairness benefit versus full model \\
\botrule
\end{tabular}%
}
\end{table*}

\subsection{Experiment 7: Calibration and Trust Reliability}\label{subsec:exp7}

The seventh experiment evaluates whether Equi-Agent improves confidence reliability. We report global expected calibration error, subgroup expected calibration error, Brier score, negative log-likelihood, and risk-coverage behavior. For risk-coverage analysis, predictions are ranked by confidence and progressively abstained from lowest-confidence to highest-confidence cases. A reliable trust-calibrated system should reduce error among retained cases while maintaining subgroup coverage.

\begin{table*}[t]
\centering
\caption{Calibration and trust reliability. Subgroup ECE reports calibration error averaged across demographic subgroups or measured in the worst-calibrated subgroup. Current results use Harvard-FairVision test predictions. Entries marked \tbd indicate calibration or subgroup files not available in the local manuscript snapshot.}
\label{tab:calibration}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}llllllll@{}}
\toprule
Method & F1 $\uparrow$ & Global ECE $\downarrow$ & Subgroup ECE $\downarrow$ & Brier $\downarrow$ & NLL $\downarrow$ & Coverage at target risk $\uparrow$ & Worst-group coverage $\uparrow$ \\
\midrule
Dynamic global prior by validation ranking & 0.473 & 0.184 & 0.636 & \tbd & \tbd & \tbd & \tbd \\
Dynamic global prior by ECE & 0.476 & 0.060 & 0.311 & \tbd & \tbd & \tbd & \tbd \\
Best standalone foundation model & 0.678 & 0.054 & 0.183 & 0.117 & 0.366 & \tbd & \tbd \\
Confidence-weighted ensemble & 0.476 & 0.077 & 0.369 & \tbd & \tbd & \tbd & \tbd \\
Mean probability ensemble & 0.469 & 0.081 & 0.311 & \tbd & \tbd & \tbd & \tbd \\
RETFound OCT & 0.678 & 0.054 & 0.183 & 0.117 & 0.366 & \tbd & \tbd \\
MIRAGE SLO & 0.538 & 0.103 & 0.320 & 0.166 & 0.514 & \tbd & \tbd \\
FLAIR SLO & 0.461 & 0.247 & 0.623 & \tbd & \tbd & \tbd & \tbd \\
RET-CLIP SLO & 0.516 & 0.245 & 0.599 & \tbd & \tbd & \tbd & \tbd \\
RetiZero SLO & 0.526 & 0.156 & 0.346 & \tbd & \tbd & \tbd & \tbd \\
UrFound OCT & 0.642 & 0.107 & 0.233 & \tbd & \tbd & \tbd & \tbd \\
UrFound SLO & 0.515 & 0.162 & 0.384 & \tbd & \tbd & \tbd & \tbd \\
VisionFM OCT & 0.617 & 0.137 & 0.272 & \tbd & \tbd & \tbd & \tbd \\
VisionFM SLO & 0.528 & 0.175 & 0.315 & \tbd & \tbd & \tbd & \tbd \\
Conformal prediction / abstention & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
\textbf{Equi-Agent} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} \\
\botrule
\end{tabular}%
}
\end{table*}

\subsection{Experiment 8: Longitudinal Glaucoma Progression Prediction}\label{subsec:exp8}

The eighth experiment evaluates whether Equi-Agent extends from cross-sectional diagnosis to longitudinal glaucoma progression prediction. In this setting, the arbitration process incorporates imaging evidence from multiple visits, temporal changes in model predictions, and clinical plausibility of progression. The longitudinal agent evaluates whether observed changes are consistent with progressive glaucomatous damage or with noise, image quality variation, or inconsistent model behavior.

We compare Equi-Agent against single-visit models, temporal models, static ensembles, and fairness-aware glaucoma progression baselines. If Equi-Agent does not achieve state-of-the-art aggregate performance, we report it as a competitive progression model with improved subgroup reliability rather than as a state-of-the-art model.

\begin{table*}[t]
\centering
\caption{Harvard-GDP glaucoma detection using OCT-compatible foundation-model encoders. RETFound OCT, VisionFM OCT, and UrFound OCT are evaluated only because their pretrained encoders can consume OCT B-scan evidence. SLO/fundus-only models are excluded from Harvard-GDP unless matching SLO/fundus inputs are available. Each row trains a GDP-specific linear probe on the official training split and reports held-out test performance.}
\label{tab:gdp_glaucoma_detection_foundation}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}llllllllll@{}}
\toprule
Method & Input & Train/Test split & F1 $\uparrow$ & Balanced Acc. $\uparrow$ & Sens. $\uparrow$ & Spec. $\uparrow$ & ECE $\downarrow$ & Worst-group F1 $\uparrow$ & $\Delta$FPR / $\Delta$FNR $\downarrow$ \\
\midrule
RETFound OCT linear probe & OCT B-scan & 600 / 400 & 0.667 & 0.701 & 0.659 & 0.742 & 0.162 & 0.529 & 0.355 / 0.291 \\
VisionFM OCT linear probe & OCT B-scan & 600 / 400 & 0.695 & 0.701 & 0.782 & 0.620 & 0.206 & 0.515 & 0.133 / 0.362 \\
UrFound OCT linear probe & OCT B-scan & 600 / 400 & 0.630 & 0.650 & 0.670 & 0.629 & 0.193 & 0.500 & 0.260 / 0.355 \\
\botrule
\end{tabular}%
}
\end{table*}

\begin{table*}[t]
\centering
\caption{Longitudinal glaucoma progression prediction on Harvard-GDP. Logistic baselines and OCT-compatible foundation-model probes use the official progression forecasting split with 300 training and 200 test examples. Temporal context tests whether dynamic arbitration generalizes beyond single-image diagnosis.}
\label{tab:longitudinal_glaucoma}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}llllllllll@{}}
\toprule
Method & Input & Temporal modeling & Fairness component & F1 $\uparrow$ & Sens. $\uparrow$ & Spec. $\uparrow$ & Worst-group F1 $\uparrow$ & $\Delta$FPR $\downarrow$ & $\Delta$FNR $\downarrow$ \\
\midrule
RNFLT logistic baseline & RNFLT & No & No & 0.206 & 0.556 & 0.621 & 0.000 & 1.000 & 1.000 \\
B-scan logistic baseline & OCT B-scan & No & No & 0.241 & 0.389 & 0.819 & 0.000 & 0.500 & 1.000 \\
Clinical tabular logistic baseline & Metadata + visual field & No & No & 0.211 & 0.222 & 0.912 & 0.000 & 0.667 & 1.000 \\
RNFLT + clinical logistic baseline & RNFLT + metadata + visual field & No & No & 0.195 & 0.222 & 0.896 & 0.000 & 0.360 & 1.000 \\
B-scan + clinical logistic baseline & OCT B-scan + metadata + visual field & No & No & 0.256 & 0.278 & 0.912 & 0.000 & 0.400 & 1.000 \\
OCT + RNFLT + clinical logistic baseline & OCT + RNFLT + metadata + visual field & No & No & 0.213 & 0.278 & 0.868 & 0.000 & 0.363 & 1.000 \\
RETFound OCT linear probe & OCT B-scan & No & No & 0.103 & 0.111 & 0.896 & 0.000 & 1.000 & 1.000 \\
VisionFM OCT linear probe & OCT B-scan & No & No & 0.167 & 0.278 & 0.797 & 0.000 & 1.000 & 1.000 \\
UrFound OCT linear probe & OCT B-scan & No & No & 0.095 & 0.111 & 0.879 & 0.000 & 1.000 & 1.000 \\
Single-visit 3D CNN & OCT & No & No & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
Single-visit 3D ViT / Swin & OCT & No & No & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
Temporal LSTM & Visit sequence & Yes & No & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
Temporal Transformer & Visit sequence & Yes & No & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
FairDist / EqEffNet-style baseline & OCT & Optional & Yes & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
Static ensemble & OCT + model outputs & Weak & No & \tbd & \tbd & \tbd & \tbd & \tbd & \tbd \\
\textbf{Equi-Agent longitudinal} & \textbf{OCT + metadata + prior visits} & \textbf{Yes} & \textbf{Yes} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} & \textbf{\tbd} \\
\botrule
\end{tabular}%
}
\end{table*}

\subsection{Experiment 9: Clinical Morphology and Error-Mode Analysis}\label{subsec:exp9}

The ninth experiment evaluates whether Equi-Agent corrects clinically interpretable failure modes. We focus on cases where base models may overcall or undercall disease because of morphology, image quality, or temporal ambiguity. For glaucoma, examples include large physiologic cup, borderline retinal nerve fiber layer thinning, poor-quality OCT scans, and subtle longitudinal progression. For each case type, we report whether Equi-Agent changes the model trust allocation, adjusts confidence, abstains, or changes the final prediction.

\begin{table*}[t]
\centering
\caption{Clinical morphology and error-mode analysis. These analyses test whether Equi-Agent corrects clinically interpretable model failure modes rather than only improving aggregate metrics.}
\label{tab:morphology_error_modes}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lllllll@{}}
\toprule
Case type & Disease task & Common baseline error & Equi-Agent mechanism & Primary metric & Secondary metric & Qualitative figure \\
\midrule
Large physiologic cup & Glaucoma & False-positive glaucoma prediction & Down-weight overreliance on cup-to-disc ratio & FPR reduction & Confidence calibration & Fig.~\tbd \\
Borderline RNFL thinning & Glaucoma & Overconfident positive prediction & Increase uncertainty or request temporal evidence & ECE reduction & Abstention accuracy & Fig.~\tbd \\
Subtle early progression & Glaucoma progression & False-negative progression prediction & Use longitudinal change and model disagreement & FNR reduction & Sensitivity gain & Fig.~\tbd \\
Poor image quality & AMD / DR / glaucoma & Overconfident incorrect prediction & Lower trust in affected model output & Brier score & Risk-coverage & Fig.~\tbd \\
Conflicting model predictions & All tasks & Unstable ensemble output & Arbitration based on priors and visual evidence & Worst-group F1 & $\Delta$FPR / $\Delta$FNR & Fig.~\tbd \\
\botrule
\end{tabular}%
}
\end{table*}

\subsection{Experiment 10: Reproducibility, Robustness, and Cost}\label{subsec:exp10}

The final experiment evaluates the reproducibility and practicality of Equi-Agent. Because agent-based systems may be sensitive to prompt wording, random seeds, and language-model configuration, we test repeated runs, prompt paraphrases, alternative LLM backbones, missing metadata, noisy priors, and external validation where available. We also report runtime, number of model calls, and average cost per case.

\begin{table*}[t]
\centering
\caption{Reproducibility, robustness, and cost analysis for Equi-Agent. Prediction agreement is measured across repeated runs or prompt variants.}
\label{tab:reproducibility}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}llllll@{}}
\toprule
Stress test & Setting & Prediction agreement $\uparrow$ & F1 change & Fairness change & Cost / latency \\
\midrule
Repeated runs & Same prompt, temperature 0 & \tbd & \tbd & \tbd & \tbd \\
Prompt paraphrase & Semantically equivalent prompts & \tbd & \tbd & \tbd & \tbd \\
Alternative LLM backbone & Replace arbitration LLM & \tbd & \tbd & \tbd & \tbd \\
Missing metadata & Remove one or more metadata fields & \tbd & \tbd & \tbd & \tbd \\
Noisy subgroup priors & Add perturbation to validation-derived priors & \tbd & \tbd & \tbd & \tbd \\
External validation & Held-out external cohort & \tbd & \tbd & \tbd & \tbd \\
Runtime analysis & Full Equi-Agent inference & N/A & N/A & N/A & \tbd \\
\botrule
\end{tabular}%
}
\end{table*}

\subsection{Statistical Analysis}\label{subsec:statistical_analysis}

All metrics are reported with 95\% confidence intervals estimated by nonparametric bootstrapping over patients. When both eyes from the same patient are present, bootstrapping is performed at the patient level to avoid treating correlated eyes as independent samples. Comparisons between models are performed using paired bootstrap tests on the same test cases. For AUROC comparisons, DeLong's test may be used as a supplementary analysis. For subgroup metrics, we report subgroup sample sizes and confidence intervals to avoid overinterpreting unstable estimates from small groups.

Primary statistical comparisons are made between Equi-Agent and the strongest validation-selected non-agentic baseline. Secondary comparisons are made against the best standalone foundation model, the best static ensemble, and the best fairness-aware baseline. Multiple comparisons are controlled using the Benjamini--Hochberg procedure when reporting large numbers of subgroup-level tests.

\subsection{Planned Acceptance Criteria}\label{subsec:acceptance_criteria}

We define success as an improvement in worst-group reliability without unacceptable loss of aggregate diagnostic performance. Specifically, Equi-Agent is considered successful if it reduces false-positive or false-negative disparities relative to the strongest non-agentic baseline while maintaining comparable or improved F1-score, balanced accuracy, and calibration. For longitudinal glaucoma progression, Equi-Agent is considered successful if it improves subgroup reliability and achieves competitive temporal prediction performance. Claims of state-of-the-art performance are made only if Equi-Agent outperforms the strongest reproduced longitudinal baselines under the same test split and evaluation protocol.

\section{Results}
\subsection{Diagnostic Performance Across Disease Tasks}
\label{subsec:results_diagnostic}
Standalone foundation-model performance varied substantially by disease task and modality. In the currently available held-out FairVision outputs, OCT-compatible encoders performed strongest overall. RETFound OCT produced the highest macro F1-score among the completed foundation-model runs (0.678), with strong AMD performance (F1 0.775, balanced accuracy 0.825), DR performance (F1 0.537, balanced accuracy 0.838), and glaucoma performance (F1 0.721, balanced accuracy 0.728). UrFound OCT produced the next-highest macro F1-score (0.642), followed by VisionFM OCT (0.617). VisionFM OCT was strong for AMD (F1 0.743, balanced accuracy 0.801) and DR (F1 0.440, balanced accuracy 0.804), while UrFound OCT achieved similar AMD and DR ranking performance and high macro specificity. MIRAGE SLO reached a macro F1-score of 0.538, with useful SLO/fundus signal for AMD and glaucoma but weaker thresholded DR performance. Other SLO/fundus encoders were weaker overall but still provided task-specific signal: VisionFM SLO, RetiZero SLO, and RET-CLIP SLO reached macro F1-scores of 0.528, 0.526, and 0.516, respectively. FLAIR SLO was weakest among the completed foundation-model runs (macro F1 0.461), largely because of lower specificity on AMD and DR.

These completed foundation-model runs show the key experimental pattern: model reliability is strongly task- and modality-dependent, and no single standalone foundation model is uniformly strongest across AMD, DR, and glaucoma. RETFound OCT provides the strongest overall diagnostic baseline, but its subgroup-disparity rows still show nontrivial intersectional false-negative and false-positive gaps, motivating arbitration beyond average test performance.

\subsection{Interim Live Equi-Agent Subset}
\label{subsec:results_interim_live_subset}
An interim live Equi-Agent run on a stratified 1,500-case FairVision subset is included as a placeholder to show the current arbitration output format (Table~\ref{tab:equi_agent_stratified1500_placeholder}). This run used 500 task-cases per disease with balanced labels and completed without API errors. Thresholded performance was strongest for glaucoma (F1 0.785), weaker for AMD (F1 0.486), and weakest for DR (F1 0.404). On the exact same 1,500 cases, Equi-Agent showed a more conservative operating point than several high-recall standalone baselines, while static fusion baselines were more specific but missed substantially more positives (Table~\ref{tab:equi_agent_stratified1500_comparison}). Because this subset is deliberately balanced and smaller than the full FairVision test set, these results are not used for final claims, but they provide an initial live-system operating point for manuscript development.

\subsection{Subgroup Fairness and Error Disparity Reduction}
\label{subsec:results_fairness}
Standalone results revealed clinically meaningful reliability heterogeneity across disease tasks and modalities. The synchronized subgroup-disparity summaries show substantial intersectional gaps despite strong aggregate RETFound performance. For example, RETFound OCT achieved the strongest macro F1-score but still showed maximum subgroup false-negative and false-positive gaps above 0.49 in glaucoma and AMD. The aggregate operating points nevertheless show why subgroup auditing is necessary: several SLO/fundus encoders perform reasonably on AMD and glaucoma but remain weak on DR, while OCT-compatible encoders perform better on AMD and DR but are not uniformly dominant across all tasks.

These findings support the premise that average performance is insufficient for model selection. A model can be preferable for one disease and subgroup but unreliable for another. Equi-Agent directly targets this failure mode by using subgroup priors and case-specific disagreement to adjust model trust.

\subsection{Calibration and Trust Reliability}
\label{subsec:results_calibration}
Calibration results showed that lower aggregate error did not necessarily correspond to better subgroup calibration. Confidence-weighted ensembling had the lowest global ECE among static baselines, while dynamic global routing by ECE further reduced global ECE. However, subgroup ECE remained high for several methods, especially when performance was dominated by a disease task with highly imbalanced threshold behavior.

Calibration varied sharply across foundation models. Among the completed runs, RETFound OCT had the lowest macro global ECE (0.054), followed by MIRAGE SLO (0.103), UrFound OCT (0.107), VisionFM OCT (0.137), RetiZero SLO (0.156), UrFound SLO (0.162), and VisionFM SLO (0.175). RET-CLIP SLO and FLAIR SLO had higher macro ECE values (0.245 and 0.247), driven by weaker calibration on AMD and DR. Subgroup ECE was often much higher than global ECE, especially for SLO/fundus encoders on AMD and DR. These calibration patterns indicate that output probabilities should not be interpreted as uniformly reliable across diseases or modalities. Equi-Agent uses these calibration failures as inputs to trust adjustment rather than hiding them behind a single aggregate score.

\subsection{Longitudinal Glaucoma Progression Prediction}
\label{subsec:results_longitudinal}
Harvard-GDP logistic baselines using RNFLT, B-scan, clinical metadata, and combined feature sets produced modest AUROC values and weak F1-scores for progression forecasting, with worst-group F1 frequently equal to zero. This indicates that longitudinal progression prediction is substantially more difficult than cross-sectional FairVision diagnosis under the current feature representation.

OCT-compatible foundation-model encoders were then evaluated on Harvard-GDP using GDP-specific linear probes. For glaucoma detection, VisionFM OCT achieved the strongest AUROC among the three OCT encoders tested (AUROC 0.796, F1 0.695), followed by RETFound OCT (AUROC 0.779, F1 0.667) and UrFound OCT (AUROC 0.687, F1 0.630). These results suggest that OCT foundation-model representations carry useful cross-sectional glaucoma signal on GDP, although subgroup error gaps remain nontrivial.

Progression forecasting remained substantially harder. VisionFM OCT showed the strongest ranking performance among the OCT foundation-model probes (AUROC 0.649), but thresholded F1 remained low (0.167) with high false-negative rate. RETFound OCT and UrFound OCT were weaker for progression under the current single-center-B-scan probe setup. These findings motivate temporal modeling and arbitration over longitudinal evidence rather than relying on a single visit or single B-scan representation.

The longitudinal extension evaluates whether OCT foundation-model features and temporal arbitration improve progression sensitivity and subgroup reliability over these logistic and single-visit foundation-model baselines. This comparison is important because a clinically useful glaucoma progression system must detect worsening disease without amplifying subgroup-specific under-detection.

\subsection{Ablation Studies}
\label{subsec:results_ablation}
The ablation study isolates the contribution of each Equi-Agent component. The key comparisons remove demographic context, subgroup priors, model-disagreement reasoning, visual evidence, calibration, and multi-agent decomposition. A randomized-prior control tests whether any fairness improvement depends on meaningful validation-derived reliability estimates rather than the mere presence of demographic fields.

The most important comparison is between full Equi-Agent and deterministic subgroup-prior routing. If the full agent improves only marginally over deterministic routing, the simpler rule-based mechanism may be preferable. If the full agent improves calibration, worst-group F1, or safety escalation behavior, this supports the value of structured agentic reasoning.

\subsection{Qualitative Case Studies}
\label{subsec:results_qualitative}
Qualitative case studies focus on clinically interpretable failure modes: large physiologic cup causing false-positive glaucoma predictions, borderline RNFL thinning with overconfident model outputs, subtle early progression, poor-quality scans, and cross-model disagreement. For each case, we report the base-model probabilities, subgroup reliability prior, agent trust adjustment, final decision, and safety recommendation. These examples demonstrate mechanism and complement the quantitative evaluation.

\section{Discussion}
\label{sec:discussion}
This study develops Equi-Agent as a framework for dynamic trust calibration in retinal foundation-model deployment. The central observation is that foundation-model reliability is not fixed. It varies by disease task, modality, subgroup, and thresholding behavior. Static fusion and global calibration can improve some aggregate metrics, but they do not directly answer which model should be trusted for a particular patient context.

The multi-agent design makes this trust decision explicit. The Vision Specialist grounds probability outputs in retinal morphology, the Functional Specialist provides glaucoma-relevant visual-field interpretation, the Equity Auditor translates validation-derived subgroup priors into trust adjustments, and the Safety Agent audits uncertainty and discordance. This decomposition is useful scientifically because it makes each source of evidence inspectable. It is also useful clinically because unsafe cases can be escalated rather than forced into a confident prediction.

The FairVision foundation-model comparison demonstrates why arbitration is needed: high aggregate performance and reliable subgroup behavior do not necessarily coincide. The critical test for Equi-Agent is therefore not only whether it improves mean thresholded performance, but whether it improves the reliability of the worst-served patients while preserving diagnostic utility.

\subsection{Fairness of Foundation-Model Comparison}
\label{subsec:discussion_fairness_protocol}
The evaluation is fair in the sense of controlled downstream comparison. All foundation models use the same FairVision splits, train only on the training split, select thresholds only on the validation split, and report metrics only on the held-out test split. They are evaluated using the modality they are designed to process, and all prediction outputs are converted to the same schema before metric computation.

This does not mean every model has been optimized to its maximum possible performance. Some models may benefit from model-specific preprocessing, full fine-tuning, prompt engineering, higher input resolution, or multimodal usage. The reported results should therefore be interpreted as controlled transfer comparisons, not as definitive leaderboards of the foundation models themselves.

\subsection{Clinical and Methodological Implications}
\label{subsec:discussion_implications}
Equi-Agent supports a deployment pattern in which foundation models are treated as evidence sources rather than autonomous diagnostic authorities. A high-confidence model output should be trusted only when it is consistent with morphology, functional context, and validation reliability for the relevant subgroup. Conversely, a borderline or discordant model output can trigger threshold adjustment, uncertainty inflation, or human review.

This approach may be especially valuable in ophthalmology because multimodal evidence is naturally complementary. OCT can reveal retinal layer pathology, SLO/fundus imaging can reveal vascular and optic-disc patterns, and visual fields can reveal functional loss. An arbitration framework can exploit this complementarity while explicitly monitoring when a modality or model is unreliable.

\subsection{Limitations}
\label{subsec:limitations}
Several limitations remain. First, subgroup estimates can be unstable when positive or negative counts are small, especially for intersectional groups. Second, demographic metadata can improve reliability calibration but also introduces ethical and methodological risk if used as a disease shortcut. Metadata-only, shuffled-metadata, randomized-prior, and counterfactual tests are therefore essential. Third, LLM-based arbitration can be sensitive to prompt wording and model version, requiring repeated-run, prompt-paraphrase, and alternative-backbone analyses. Finally, longitudinal glaucoma progression requires stronger temporal modeling than the current logistic baselines provide.

\section{Conclusion}
\label{sec:conclusion}
Equi-Agent reframes retinal foundation-model deployment as a problem of dynamic trust calibration. FairVision experiments show that standalone foundation models and simple ensembles can have substantial subgroup reliability gaps, motivating contextual arbitration. The proposed multi-agent system provides an interpretable structure for combining image evidence, patient metadata, subgroup priors, functional testing, guideline context, and safety review. This framework offers a practical route toward ophthalmic AI systems that are not only accurate on average, but more reliable for the patients most vulnerable to model failure.

\section*{Data Availability}
The experiments use Harvard-FairVision and Harvard-GDP data. Dataset access, licensing conditions, and any restrictions on redistribution will be described here before submission. Derived prediction files, subgroup metrics, and manuscript tables are produced under the project outputs directory and can be shared subject to dataset-use agreements.

\section*{Code Availability}
The Equi-Agent codebase contains scripts for manifest construction, foundation-model prediction extraction, threshold tuning, subgroup-prior construction, dynamic routing, metric computation, and manuscript table generation. A reproducible release with environment files, command-line entry points, and trained lightweight probes will be prepared before submission, subject to the licensing terms of the external foundation-model repositories.

\section*{Acknowledgements}
\tbd{}

\section*{Author Contributions}
A.M.A.R. designed the study, implemented the Equi-Agent framework, ran experiments, and drafted the manuscript. Y.A., N.B., and Y.T. contributed to experimental design, analysis, interpretation, and manuscript revision. Final author roles will be updated before submission.

\section*{Competing Interests}
The authors declare no competing interests. This statement will be updated before submission if needed.

\section*{Ethics and Reporting}
All analyses use retrospective ophthalmic datasets. The final manuscript will report dataset approvals, de-identification status, and relevant data-use agreements according to the requirements of the source datasets and target journal.

\begin{appendices}
\section{Supplementary Material}\label{secA1}
Supplementary implementation details, additional subgroup tables, and robustness analyses will be included here.
\end{appendices}

\bibliography{sn-bibliography}

\end{document}
